The importance of transient resonances in extreme-mass-ratio inspirals

### 2016-11-10 referee_response.txt

We thank the referee for their careful reading of our document, especially given its length! We have implemented almost all of their suggestions, as indicated below, and we feel that the manuscript has been improved by these changes.

In addition to the changes commented below, we have updated some of the references and added the referee to the acknowledgements. We have also removed the mention in the Introduction that $\Omega_r < \Omega_\theta < \Omega_\phi$ as Maarten van de Meent has pointed out to us that this hierarchy is not always true; this does not effect anything else in the paper.

@@

#1
In Fig.2, is it possible for authors to add the values of semilatus
rectum p and eccentricity e at t=t_{res}? I would like to compare the
values with those in Fig.8 as a reference.

>> The values on resonance are p = 7.67, e = 0.674 and cos(iota) = 0.497. Figure 8 shows the orbital parameters at plunge, rather than at resonance, so these can't be compared directly. This example doesn't reach plunge during it's two-year window, so we think it would be confusing to add these numbers to the text. This example is to illustrate what resonances can be like, and the calculation of what they are typically like is left for later on in the paper.

@@

#2
In Sec.V.A the distribution of inclination angle is assume to be
uniform. Do the authors know any reference to justify this assumption?

For example, in Ref.[78], Flanagan and Hinderer (2007), Ref.[83], Sago
et al, and Ref.[84], Ganz et al, of this manuscript, it is found that
gravitational radiation reaction drives the orbital plane to
anti-align with the symmetry axis. The effect would be very small, but
it might be commented on the effect (ex. in a footnote).

>> The true inclination distribution is currently uncertain, as we are not sure of the dynamics at the very centre of galaxies. We used a uniform distribution as the least constraining proposal in absence of better information. It is true that radiation reaction will lead to anti-alignment, but we are considering the initial distribution of orbits from which inspirals begin; this is the same reason why not all the orbits are circular, despite the circularising effect of radiation reaction. Adding a comment on potential alignment is a sensible recommendation, so we have added a footnote to explain this while discussing the evolution of the analytic kludge model.

@@

#3
In the last paragraph in Sec.VI, there is the sentence ends with "if
the phase on resonance and the magnitude of the self-force can be
predicted with sufficient accuracy.".

If the authors would think that interpolation formulas of the
gravitational self-force in Kerr spacetime, which might be an
extension of Ref.[75], Warburton et al (2012), could be used for this
purpose, it might be commented in the same paragraph in Sec.VI.

>> We think that interpolation formulae could certainly be used if there is sufficient numerical data to calibrate these. We have added a footnote to highlight this as suggested.

### 2016-11-10 res_char.tex
Referee comments

Flanagan and Hinderer~\cite{Flanagan2012} highlighted a previously overlooked phenomenon that occurs in the general relativistic two-body problem, that of transient resonances. Geodesic orbits in GR have three associated frequencies: the radial frequency $\Omega_r$, the polar frequency $\Omega_\theta$ and the azimuthal frequency $\Omega_\phi$.\footnote{In the strong-field regime, it is possible to have isofrequency pairings, where two different orbits share the same orbital frequencies~\cite{Warburton2013}. The evolution of the frequencies still differ, such that orbital trajectories can be reconstructed.} The first two describe libration and the third rotation (except in the case of polar orbits where $\Omega_\theta$ also describes rotation)~\cite{Goldstein2002}. % Section 10.6
In the weak-field limit, these all tend towards the Keplerian frequency; in the strong-field regime $\Omega_r < \Omega_\theta < \Omega_\phi$, and they may differ significantly. For EMRIs, the evolution timescale is much longer than the orbital period such that the motion of the smaller body is approximately geodesic over orbital timescales. The evolution inspiral of the orbit can be approximated as a series of geodesics using the osculating element formalism~\cite{Pound2008,Gair2011a}. During this evolution, the frequencies may become commensurate: resonances occur when the radial and polar frequencies are rational multiples of each other:
\begin{equation}
\nu \equiv \frac{\Omega_r}{\Omega_\theta} = \frac{n_\theta}{n_r},
\end{equation}
where $n_r$ and $n_\theta$ are integers (with no common factors). During resonance, terms in the self-force that usually average to zero can combine coherently, significantly impacting the orbital motion~\cite{Flanagan2012a}.

@@

We define the evolution timescale as
\begin{equation}
\tau\sub{ev} = \frac{\nu}{\dot{\nu}},
\end{equation}
where an overdot denotes a derivative with respect to $t$. In general, away from resonance, we take $\nu \equiv \Omega_r/\Omega_\theta$. This timescale sets the period over which there is a significant change in the frequencies. It acts as an inspiral timescale. It is long in all cases we study, $\tau\sub{ev} \sim \order{T/\eta}$. It is this property which makes EMRIs interesting, as we can follow the waveform for many cycles, accruing high signal-to-noise ratios (SNRs). This is also what allows us to use the adiabatic prescription, as it means the trajectory moves slowly through different orbital parameters.

@@

To evolve the orbits, we start with orbital periods of $T_0$ and then use the analytic kludge (AK) prescription of Barack and Cutler~\cite{Barack2004}. This is similar to the NK approach, but uses a series of Keplerian ellipses rather than Kerr geodesics. Evolution of the orbit includes the effects of periapse precession, Lense--Thirring
precession, and radiation reaction calculated using PN expressions. AK waveforms are less computationally expensive than NK waveforms, allowing us to simulate a large population of EMRIs. We follow inspirals until the last stable orbit (LSO).\footnote{The LSO is determined numerically by calculating the roots of $V_r(r) = 0$, which we denote in ascending order by $r_4 \leq r_3 \leq r\sub{p} \leq r\sub{a}$, and stopping the evolution when $r_3 = r_p$, which designates the orbit as marginally stable. This ignores the (small) influence of the self-force~\cite{Isoyama2014}.} Each of these systems is then evolved backwards for some time $t\sub{insp}$, chosen uniformly from the range $[0,t\sub{life}]$ for a mission lifetime $t\sub{life}$, and the expected GWs are calculated using the AK formalism.

@@

Adiabatic waveform models can still be used for EMRIs away from resonance. Therefore it may be possible to stitch together waveforms from a sequence of adiabatic evolutions, if the phase on resonance and the magnitude of the self-force can be predicted with sufficient accuracy. Such hybrid models merit further study as relatively simple ways of incorporating resonance effects into adiabatic models.

>>>>

Flanagan and Hinderer~\cite{Flanagan2012} highlighted a previously overlooked phenomenon that occurs in the general relativistic two-body problem, that of transient resonances. Geodesic orbits in GR have three associated frequencies: the radial frequency $\Omega_r$, the polar frequency $\Omega_\theta$ and the azimuthal frequency $\Omega_\phi$.\footnote{In the strong-field regime, it is possible to have isofrequency pairings, where two different orbits share the same orbital frequencies~\cite{Warburton2013}. The evolution of the frequencies still differ, such that orbital trajectories can be reconstructed.} The first two describe libration and the third rotation (except in the case of polar orbits where $\Omega_\theta$ also describes rotation)~\cite{Goldstein2002}. % Section 10.6
In the weak-field limit, these all tend towards the Keplerian frequency; in the strong-field regime they may differ significantly. For EMRIs, the evolution timescale is much longer than the orbital period such that the motion of the smaller body is approximately geodesic over orbital timescales. The evolution inspiral of the orbit can be approximated as a series of geodesics using the osculating element formalism~\cite{Pound2008,Gair2011a}. During this evolution, the frequencies may become commensurate: resonances occur when the radial and polar frequencies are rational multiples of each other:
\begin{equation}
\nu \equiv \frac{\Omega_r}{\Omega_\theta} = \frac{n_\theta}{n_r},
\end{equation}
where $n_r$ and $n_\theta$ are integers (with no common factors). During resonance, terms in the self-force that usually average to zero can combine coherently, significantly impacting the orbital motion~\cite{Flanagan2012a}.

@@

We define the evolution timescale as
\begin{equation}
\tau\sub{ev} = \frac{\nu}{\dot{\nu}},
\end{equation}
where an overdot denotes a derivative with respect to $t$. In general, away from resonance, we take $\nu \equiv \Omega_r/\Omega_\theta < 1$. This timescale sets the period over which there is a significant change in the frequencies. It acts as an inspiral timescale. It is long in all cases we study, $\tau\sub{ev} \sim \order{T/\eta}$. It is this property which makes EMRIs interesting, as we can follow the waveform for many cycles, accruing high signal-to-noise ratios (SNRs). This is also what allows us to use the adiabatic prescription, as it means the trajectory moves slowly through different orbital parameters.

@@

To evolve the orbits, we start with orbital periods of $T_0$ and then use the analytic kludge (AK) prescription of Barack and Cutler~\cite{Barack2004}. This is similar to the NK approach, but uses a series of Keplerian ellipses rather than Kerr geodesics. Evolution of the orbit includes the effects of periapse precession, Lense--Thirring
precession, and radiation reaction calculated using PN expressions.\footnote{While the AK approach does include these relativistic effects, it does not capture the full nature of the evolution: for example, it assumes that the angle between the orbital angular momentum and the SMBH spin is constant~\cite{Barack2004}, whereas radiation reaction should push the orbital plane towards being anitaligned~\cite{Flanagan2007}. The approximate nature of the AK evolution should not effect our results more than the uncertainty in the initial conditions for EMRI orbits.} AK waveforms are less computationally expensive than NK waveforms, allowing us to simulate a large population of EMRIs. We follow inspirals until the last stable orbit (LSO).\footnote{The LSO is determined numerically by calculating the roots of $V_r(r) = 0$, which we denote in ascending order by $r_4 \leq r_3 \leq r\sub{p} \leq r\sub{a}$, and stopping the evolution when $r_3 = r_p$, which designates the orbit as marginally stable. This ignores the (small) influence of the self-force~\cite{Isoyama2014}.} Each of these systems is then evolved backwards for some time $t\sub{insp}$, chosen uniformly from the range $[0,t\sub{life}]$ for a mission lifetime $t\sub{life}$, and the expected GWs are calculated using the AK formalism.

@@

Adiabatic waveform models can still be used for EMRIs away from resonance. Therefore, it may be possible to stitch together waveforms from a sequence of adiabatic evolutions, if the phase on resonance and the magnitude of the self-force can be predicted with sufficient accuracy.\footnote{This may be possible using interpolation schemes~\cite{Warburton2012} if there is sufficient numerical data available for calibration.} Such hybrid models merit further study as relatively simple ways of incorporating resonance effects into adiabatic models.
